Bayesian Detection of Bias in Peremptory Challenges Using Historical Strike Data

### 2023-01-20 analysis/paper_draft3.Rmd
minor edits; add ref to Carvalho2021

The use of the power prior in this context allows lawyers to explicitly control the assumed degree of incompatibility between the current and historical trial data. A commonly adopted practice is to fix the power prior parameter $\alpha$ and assess the sensitivity of results to different values [@Carvalho & Ibrahim]. Specifically, lower values (closer to 0) of $\alpha$ should be selected if incompatibility is expected to be high. Higher values (closer to 1) of the power prior parameter should be selected if incompatibility is expected to be low. Accordingly, the lawyer bringing the _Batson_ challenge, after choosing the value of $\alpha$, must be able to defend that choice as appropriately corresponding to the likely degree of incompatibility between the current and historical strike data in the particular case. Since this overall procedure is to estimate bias to support the bringing of a _Batson_ challenge, the senstivity of bias detection to the value of $\alpha$ is used to evaluate sensitivity of the estimation of bias, not to indicate whether a _Batson_ challenge should succeed.

@@

To generate data, we generated strike data for a current trial using seven different values of the bias parameter $b_{curr}=\{0,0.5,1,1.5,2, 2.5,3\}$. To evaluate the strength of the power prior as a function of the degree of incompatibility between current and historical strike data ($\lvert b_{curr} - b_{hist} \rvert > 0$), we assigned $b_{hist}$ the same bias values as $b_{curr}$, i.e., $b_{hist}=\{0,0.5,1,1.5,2, 2.5,3\}$. We did this to simulate noise in the historical strike data that can arise due to not-at-random missingness or because a confounder (e.g. charge severity) affects $b_{curr}$ but not of all the past trials in the historical strike data.

For our simulation, we define bias detection as the proportion of times we identify bias when bias is present in the current trial, i.e., $b_{curr} > 0$ , based on whether credible intervals of the bias parameter exclude zero. Since we use a weakly-informative prior that pulls the estimate of $b_{curr}$ towards zero, we focus on how accurately our model can detect that bias, _not_ on how accurately the model can recover the true value of $b_{curr}$ (i.e., a traditional coverage rate). For instance, a 90\% bias detection rate means that in 900 of 1000 model fits, a given credible interval does not contain zero if the $b_{curr}$ is not zero. Put another way, if we set $b_{curr}>0$, we calculate the proportion of the 90\% credible intervals that lie to the right of zero (lower bound is positive) among the 1000 model fits. For each scenario, we generated 1000 data sets and fit the model on those data sets. For each replicate, we generate a MCMC sample of 10,000 iterations with a burn-in period of 1000 iterations. For each replicate, we calculate the posterior mean and $80\%$, $90\%$, and $95\%$ highest posterior density intervals for $b_{curr}$.

@@

For the prototype app, we set two cognizable-class options: race and gender. This was because the standard juror questionnaire in this court did not ask about other possible cognizable classes. The data also did not expressly indicate the race or gender of the criminal defendant(s).

In the default setting, the pull-down menus for prosecutor and defense are set to "None". As a result, the app ignores any historical strike data and estimates the posterior distributions of the bias parameter for prosecutor and defense based only on the strike tally data and the initial prior ($b \sim N(0,2)$). Historical data will be included if the name(s) of a prosection or defense attorney are selected from the pull-down menu. Subsequently, the amount of historical data to be borrowed can be selected by the choice of $\alpha$.

@@

Statistical methods for _Batson_ and similar challenges aim to help measure how much prospective jurors' race, gender, or other cognizable-class membership affected an attorney's use of peremptory strikes against them. These statistical methods do not aim to  _prove_ bias alone, but rather to lend support to an inference of bias and a subsequent _Batson_ challenge. Previously used methods have either ignored historical trial data, or conducted analyses where current and historical data are merged, with no regard for possible incompatibility between trials. The Bayesian approach presented here accounts for available historical strike data when estimating the posterior for the bias parameter. By incorporating the power prior, one can assess how sensitive that posterior estimate is to the available historical strike data. This is appealing because it makes assumptions about the incompatibility between current and historical trial data an explicit part of the process of identifying bias and the decision of whether to bring a _Batson_ challenge or not.

Moreover, with a tool like our prototype app, attorneys can use the approach presented here to better detect strike bias for _Batson_ and related challenges. The posterior distribution of the striking attorney's bias parameter can help the lawyer decide in real time whether to make a  _Batson_ challenge at all. If argued, the attorney can introduce that posterior distribution as relevant evidence for presenting a prima facie case and of ultimately proving illegal bias (Figure \@ref(fig:batsonworkflowfig)).

@@

US law currently affords a way to ameliorate this missing data problem. The public has a presumptive First Amendment right to attend jury voir dire in criminal cases, even if prosecutor and defendant agree to exclude the public. Similarly, a criminal defendant has a Sixth Amendment right to a public trial that presumptively precludes excluding the public from the voir dire of prospective jurors [@press1984; @presley2010]. If these rights are asserted, any person can often attend jury voir dire and collect relevant strike data based on what they observe in real time.  Thus, though historical strike data is hard to collect retrospectively, such courtroom observers can prospectively build up a dataset for strike data that they would have been less likely to obtain from court records after the fact.

@@

By their choice of $\alpha$, the lawyer explicitly and transparently denotes the degree of weight placed on the historical strike data in the analysis. This transparency matters, because it thwarts lawyer efforts to strategically select $\alpha$ values to prove or disprove illegal strike bias. Rather, at the key steps of the _Batson_ challenge process (Figure \@ref(fig:batsonworkflowfig)), the lawyers on both sides of a _Batson_ challenge know that the lawyers on the other side can always directly and repeatedly assess how sensitive any bias estimation is to the choice of $\alpha$. In any case, the choice of $\alpha$ cannot always determine whether a trial judge finds illegal bias, because no approach to estimating bias based on historical strike data has that ability.

>>>>

The use of the power prior in this context allows lawyers to explicitly control the assumed degree of incompatibility between the current and historical trial data. A commonly adopted practice is to fix the power prior parameter $\alpha$ and assess the sensitivity of results to different values [@Carvalho2021, 5252; @Ibrahim2015, 3734]. Specifically, lower values (closer to 0) of $\alpha$ should be selected if incompatibility is expected to be high. Higher values (closer to 1) of the power prior parameter should be selected if incompatibility is expected to be low. Accordingly, the lawyer bringing the _Batson_ challenge, after choosing the value of $\alpha$, must be able to defend that choice as appropriately corresponding to the likely degree of incompatibility between the current and historical strike data in the particular case. Since this overall procedure is to estimate bias to support the bringing of a _Batson_ challenge, the senstivity of bias detection to the value of $\alpha$ is used to evaluate sensitivity of the estimation of bias, not to indicate whether a _Batson_ challenge should succeed.

@@

To generate data, we generated strike data for a current trial using seven different values of the bias parameter $b_{curr}=\{0,0.5,1,1.5,2, 2.5,3\}$. To evaluate the strength of the power prior as a function of the degree of incompatibility between current and historical strike data ($\lvert b_{curr} - b_{hist} \rvert > 0$), we assigned $b_{hist}$ the same bias values as $b_{curr}$, i.e., $b_{hist}=\{0,0.5,1,1.5,2, 2.5,3\}$. We did this to simulate noise in the historical strike data that can arise due to not-at-random missingness or because a confounder (e.g. defendant race, charge severity) affects $b_{curr}$ but not of all the past trials in the historical strike data.

For our simulation, we define bias detection as the proportion of times we identify bias when bias is present in the current trial, i.e., $b_{curr} > 0$ , based on whether credible intervals of the bias parameter exclude zero. Since we use a weakly-informative prior that pulls the estimate of $b_{curr}$ towards zero, we focus on how accurately our model can detect that bias, _not_ on how accurately the model can recover the true value of $b_{curr}$ (i.e., a traditional coverage rate). For instance, a 90\% bias detection rate means that in 900 of 1000 model fits, a given credible interval does not contain zero if the $b_{curr}$ is not zero. Put another way, if we set $b_{curr}>0$, we calculate the proportion of the 90\% credible intervals that lie to the right of zero (lower bound is positive) among the 1000 model fits. For each scenario, we generated 1000 data sets and fit the model on those data sets. For each replicate, we generate a MCMC sample of 10,000 iterations with a burn-in period of 1000 iterations. For each replicate, we calculated the posterior mean and $80\%$, $90\%$, and $95\%$ highest posterior density intervals for $b_{curr}$.

@@

For the prototype app, we set two cognizable-class options: race and gender. This was because the standard juror questionnaire in this court did not ask about other possible cognizable classes. The data also did not expressly indicate the race or gender of the criminal defendant(s). For the levels of race and gender, we were limited by choices in the juror questionnaire (e.g., for gender, "male" and "female").

In the default setting, the pull-down menus for prosecutor and defense are set to "None". As a result, the app ignores any historical strike data and estimates the posterior distributions of the bias parameter for prosecutor and defense based only on the strike tally data and the initial prior ($b \sim N(0,2)$). Historical data will be included if the app user selects the name(s) of a prosection or defense attorney from the pull-down menu. Subsequently, the amount of historical data to be borrowed can be selected by the choice of $\alpha$.

@@

Statistical methods for _Batson_ and similar challenges aim to help measure how much prospective jurors' race, gender, or other cognizable-class membership affected an attorney's use of peremptory strikes against them. These statistical methods do not aim to  prove illegal bias alone, but rather to lend support to an inference of bias and a subsequent _Batson_ challenge. The Bayesian approach presented here accounts for available historical strike data when estimating the posterior for the bias parameter. By incorporating the power prior, one can assess how sensitive that posterior estimate is to the available historical strike data. This is appealing because it makes assumptions about the incompatibility between current and historical trial data an explicit part of the process of identifying bias and the decision of whether to bring a _Batson_ challenge or not.

Moreover, with a tool like our prototype app, attorneys can use the approach presented here to better detect strike bias for purposes of bringing or satisfying proof burden at different stages of _Batson_ and related challenges. The posterior distribution of the striking attorney's bias parameter can help the lawyer decide in real time whether to make a  _Batson_ challenge at all. If argued, the attorney can introduce that posterior distribution as relevant evidence, along with any other admissible evidence, for presenting a prima facie case and of ultimately proving illegal bias (Figure \@ref(fig:batsonworkflowfig)).

@@

US law currently affords a way to reduce this missing data problem. The public has a presumptive First Amendment right to attend jury voir dire in criminal cases, even if prosecutor and defendant agree to exclude the public. Similarly, a criminal defendant has a Sixth Amendment right to a public trial that presumptively precludes excluding the public from the voir dire of prospective jurors [@press1984; @presley2010]. If these rights are asserted, any person can often attend jury voir dire and collect relevant strike data based on what they observe in real time.  Thus, though historical strike data is hard to collect retrospectively, such courtroom observers can prospectively build up a dataset for strike data that they would have been less likely to obtain from court records after the fact.

@@

By their choice of $\alpha$, the lawyer explicitly and transparently denotes the degree of weight placed on the historical strike data in the analysis. This transparency matters, because it thwarts lawyer efforts to strategically select $\alpha$ values to prove or disprove illegal strike bias. Rather, at the key steps of the _Batson_ challenge process (Figure \@ref(fig:batsonworkflowfig)), the lawyers on both sides of a _Batson_ challenge know that the lawyers on the other side can always directly and repeatedly assess how sensitive any bias estimation is to the choice of $\alpha$. In any case, the choice of $\alpha$ cannot always determine whether a trial judge finds that illegal bias is more likely than not (e.g., a _Batson_ violation), because no approach to estimating bias based on historical strike data has that ability.
